---
title: "Figure 4 – Microbial Co-occurrence Network Tutorial"
subtitle: "SPIEC-EASI networks of gut function and faecal metabolites"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width  = 8,
  fig.height = 6
)
```

## Overview

This tutorial reproduces Figure 4 of the DIME paper. We built co-occurrence networks that
capture statistical associations between gut microbiome **functional features** (PFAM
protein families, a proxy for what genes the microbiome encodes) and **faecal
metabolites** in each diet arm.

| Section | Analysis |
|---|---|
| 3 | Explore a network with `igraph` |
| 4 | Build and visualise the paired network with `NetCoMi` |
| 5 | Compare HB vs LB network topology |
| 6 | Subgraph of key metabolites |

### Why SPIEC-EASI?

Microbial abundance data are compositional (proportions sum to 1), which makes naive
Pearson correlations spurious. **SPIEC-EASI** (Sparse Inverse Covariance Estimation for
Ecological Association Inference) addresses this by fitting a sparse inverse-covariance
model on centred log-ratio transformed data, giving partial correlations that are robust
to compositional bias.

Network construction with SPIEC-EASI takes a while, so two pre-computed networks
are provided in `data/derived/networks/`. This tutorial loads those directly.

---

## 1  Load packages

```{r load}
library(tidyverse)
library(igraph)    # igraph::degree, igraph::betweenness, igraph::subgraph, …
library(NetCoMi)   # NetCoMi::netConstruct, netAnalyze, netCompare, diffnet
library(readxl)    # readxl::read_xlsx for metabolite annotations
library(ggrepel)   # ggrepel::geom_text_repel for degree-distribution labels

source("../R/data.R")    # read_metadata()
source("../R/theme.R")   # BIOACTIVE_COLORS, THEME_DIME
```

---

## 2  Load data

### 2.1 Pre-computed networks

Each network is stored in two formats:

- **igraph object** (`.igraph.Rds`) — the network as a graph, with vertex and edge
  attributes already attached (node type, edge sign, partial correlation weight)
- **Association matrix** (`.association.Rds`) — the SPIEC-EASI partial-correlation
  matrix, used to re-construct the network inside `NetCoMi`

```{r load-networks}
ig_hb  <- readRDS("../data/derived/networks/after_high.igraph.Rds")
ig_lb  <- readRDS("../data/derived/networks/after_low.igraph.Rds")

ass_hb <- readRDS("../data/derived/networks/after_high.association.Rds")
ass_lb <- readRDS("../data/derived/networks/after_low.association.Rds")

# Association matrices are data frames with an 'id' row-label column
mat_hb <- ass_hb |> as.matrix()
mat_lb <- ass_lb |> as.matrix()

cat("HB network: ", igraph::vcount(ig_hb), "vertices,",
    igraph::ecount(ig_hb), "edges\n")
cat("LB network: ", igraph::vcount(ig_lb), "vertices,",
    igraph::ecount(ig_lb), "edges\n")
```

### 2.2 Node metadata

Each vertex is either a PFAM protein family (type `"a"`) or a faecal metabolite peak
(type `"b"`). We load annotations for both and combine them into one metadata table that
will be used to colour and label nodes.

```{r load-metadata}
# Metadata: which sample is in which diet arm
tbl_md <- read_metadata("../data/source/sample_metadata.csv")

# Functional annotation for PFAM nodes
tbl_pfam_md <- read_delim(
  "../data/derived/networks/feature_md.EM.PFAML0.tsv",
  show_col_types = FALSE
) |>
  rename(node_id = 1) |>
  mutate(node_type = "PFAM")

# Metabolite annotations + differential-abundance direction
tbl_metab_ttest <- readxl::read_xlsx(
  "../data/source/untargetted_metabolomics/test_results.xlsx"
) |>
  mutate(
    node_id  = paste0(Project, `Feature ID`, `Ion mode`),
    enr_in   = if_else(`Individual Estimate` > 0, "High Bioactive", "Low Bioactive")
  ) |>
  select(node_id, enr_in)

tbl_metab_annot <- readxl::read_xlsx(
  "../data/source/untargetted_metabolomics/initial_identifications.xlsx"
) |>
  mutate(node_id = paste0(project, feature_id, ionmode)) |>
  group_by(node_id) |>
  slice_max(score, n = 1) |>
  ungroup() |>
  select(node_id, plot_manual_label)

# Combined metadata: one row per vertex that appears in the HB network
trim_lineage <- function(x) {
  parts <- strsplit(x, ";", fixed = TRUE)[[1]]
  if (length(parts) == 1) return(x)
  known <- parts[parts != "?"]
  paste(tail(known, 2), collapse = ";")   # last two known levels
}

tbl_node_md <- bind_rows(
  tbl_pfam_md |> select(node_id, node_type),
  tbl_metab_ttest |> mutate(node_type = "metabolite")
) |>
  left_join(tbl_metab_annot, by = "node_id") |>
  left_join(tbl_metab_ttest, by = c("node_id", "enr_in")) |>
  mutate(
    # Colour category: PFAM nodes one colour, metabolites coloured by diet direction
    colour_cat  = case_when(
      node_type == "PFAM"     ~ "PFAM",
      enr_in == "High Bioactive" ~ "High Bioactive",
      enr_in == "Low Bioactive"  ~ "Low Bioactive",
      TRUE                       ~ "PFAM"
    ),
    short_label = map_chr(node_id, trim_lineage)
  )
```

---

## 3  Explore the High Bioactive network with `igraph`

Before running formal analysis, it is worth exploring basic network properties. The
igraph object has vertex and edge attributes set during network construction.

### 3.1 Vertex and edge attributes

```{r igraph-attributes}
# Vertex attributes
igraph::vertex_attr_names(ig_hb)

# Edge attributes
igraph::edge_attr_names(ig_hb)

# How many nodes of each type?
table(igraph::V(ig_hb)$type)   # "a" = PFAM, "b" = metabolite
```

### 3.2 Degree distribution

**Degree** = number of neighbours a node has. Hubs (high-degree nodes) are biological
candidates for coordinating functional responses.

```{r degree-dist, fig.height=4}
deg_hb <- igraph::degree(ig_hb)

tbl_deg <- tibble(
  node_id = names(deg_hb),
  degree  = deg_hb,
  type    = igraph::V(ig_hb)$type
) |>
  mutate(node_type = if_else(type == "a", "PFAM", "Metabolite"))

ggplot(tbl_deg, aes(x = degree, fill = node_type)) +
  geom_histogram(binwidth = 1, position = "dodge") +
  labs(
    title = "Degree distribution – High Bioactive network",
    x = "Degree (number of edges)", y = "Count", fill = "Node type"
  ) +
  THEME_DIME
```

### 3.3 Betweenness centrality

**Betweenness** measures how often a node lies on the shortest path between other nodes.
High-betweenness nodes are network bridges — disrupting them would disconnect large parts
of the network.

```{r betweenness, fig.height=4}
betw_hb <- igraph::betweenness(ig_hb, normalized = TRUE)

tbl_betw <- tibble(
  node_id    = names(betw_hb),
  betweenness = betw_hb
) |>
  arrange(desc(betweenness)) |>
  head(20)

ggplot(tbl_betw, aes(x = reorder(node_id, betweenness), y = betweenness)) +
  geom_col(fill = unname(BIOACTIVE_COLORS["High Bioactive"])) +
  coord_flip() +
  labs(
    title = "Top 20 nodes by betweenness centrality (HB network)",
    x = NULL, y = "Normalised betweenness"
  ) +
  THEME_DIME
```

### 3.4 Edge sign distribution

SPIEC-EASI returns partial correlations, so edges can be positive (species that co-occur)
or negative (potential competition or mutual exclusion).

```{r edge-signs, fig.height=3, fig.width=5}
tbl_edges <- tibble(
  corr  = igraph::E(ig_hb)$corr,
  inter = igraph::E(ig_hb)$inter
) |>
  mutate(
    sign      = if_else(corr > 0, "Positive", "Negative"),
    edge_type = if_else(inter, "Inter-domain\n(PFAM–Metabolite)", "Within-domain")
  )

ggplot(tbl_edges, aes(x = corr, fill = sign)) +
  geom_histogram(bins = 40) +
  facet_wrap(~edge_type) +
  scale_fill_manual(values = c(Positive = "forestgreen", Negative = "brown2")) +
  labs(
    title = "Distribution of partial correlation coefficients (HB network)",
    x = "Partial correlation", y = "Count", fill = "Sign"
  ) +
  THEME_DIME
```

---

## 4  Paired network analysis with `NetCoMi`

`NetCoMi` provides tools specifically designed for comparing microbial co-occurrence
networks. We pass our pre-computed SPIEC-EASI association matrices directly as partial
correlations, bypassing the need to run SPIEC-EASI again.

### 4.1 Construct network objects

```{r netcomi-construct}
# netConstruct() takes the two association matrices and builds internal
# network representations for downstream analysis and plotting.
# dataType = "partialCorr" treat input as already-computed partial correlations
# sparsMethod = "none"     the SPIEC-EASI refit already selected edges
net_pair <- NetCoMi::netConstruct(
  mat_hb, mat_lb,
  dataType  = "partialCorr",
  sparsMethod = "none",
  verbose   = 0
)
```

### 4.2 Analyse network properties

`netAnalyze()` computes centrality measures (degree, betweenness, closeness, eigenvector)
and identifies **hubs** as nodes above the 95th percentile for betweenness.

```{r netcomi-analyse}
net_analysis <- NetCoMi::netAnalyze(
  net_pair,
  centrLCC    = FALSE,       # don't restrict to the largest connected component
  clustMethod = "cluster_fast_greedy",   # community detection algorithm
  hubPar      = "betweenness",           # define hubs by betweenness
  hubQuant    = 0.95,
  normDeg     = TRUE, normBetw = TRUE, normClose = TRUE, normEigen = TRUE
)

# The summary gives per-node centrality scores and global network properties
s <- summary(net_analysis,
             groupNames = c("High Bioactive", "Low Bioactive"),
             showCentr  = c("degree", "between"),
             numbNodes  = 8)
```

### 4.3 Visualise the paired network

`plot()` on a `microNetProps` object draws both networks side by side with a shared
layout so topological differences are immediately visible. Circles = PFAM nodes;
squares = metabolite nodes. Node colour indicates whether a metabolite was higher in
HB (green) or LB (blue); PFAM nodes are shown in orange.

```{r netcomi-plot, fig.width=12, fig.height=6}
# Build colour and shape vectors keyed by node name
node_colours <- setNames(
  tbl_node_md$colour_cat,
  tbl_node_md$node_id
)
node_shapes <- setNames(
  if_else(
    tbl_node_md$node_id %in% igraph::V(ig_hb)$name[igraph::V(ig_hb)$type == "a"],
    "circle", "square"
  ),
  tbl_node_md$node_id
)

BIO_COLORS <- c(
  "High Bioactive" = unname(BIOACTIVE_COLORS["High Bioactive"]),
  "Low Bioactive"  = unname(BIOACTIVE_COLORS["Low Bioactive"]),
  "PFAM"           = "#f58231"
)

plot(
  net_analysis,
  # Layout
  sameLayout  = TRUE,      # identical layout for both networks — easier to compare
  layoutGroup = "union",   # base layout on the union of both network edges
  rmSingles   = "inboth",  # drop isolated nodes that appear in neither network
  # Node appearance
  nodeSize       = "degree",
  nodeColor      = "feature",
  colorVec       = list(unname(BIO_COLORS), unname(BIO_COLORS)),
  featVecCol     = node_colours,
  nodeShape      = c("circle", "square"),
  featVecShape   = setNames(
    igraph::V(ig_hb)$type,
    igraph::V(ig_hb)$name
  ),
  nodeTransp     = 20,
  cexNodes       = 1,
  nodeSizeSpread = 1.5,
  # Labels — only show hub labels to keep the plot readable
  cexLabels      = 0,
  highlightHubs  = TRUE,
  cexHubLabels   = 0.4,
  # Edges
  edgeWidth      = 2,
  edgeTranspLow  = 50,
  edgeTranspHigh = 20,
  # Titles
  groupNames = c("High Bioactive", "Low Bioactive"),
  cexTitle   = 1.2,
  mar        = c(1, 1, 3, 1)
)
```

---

## 5  Compare HB and LB networks

`netCompare()` formally compares global network properties between the two conditions.
With `permTest = FALSE` we get estimates but not permutation-based p-values
(permutation testing is computationally expensive; set `permTest = TRUE` for publication).

```{r netcomi-compare}
net_comp <- NetCoMi::netCompare(
  net_analysis,
  permTest = FALSE,
  verbose  = FALSE,
  seed     = 123456
)

# Summary of global network properties side by side
summary(
  net_comp,
  groupNames = c("High Bioactive", "Low Bioactive"),
  showCentr  = c("degree", "between"),
  numbNodes  = 5
)
```

### Differential network — edges that change between arms

`diffnet()` identifies edges whose partial correlation differs significantly between HB
and LB using Fisher's test on the z-transformed partial correlations.

```{r diffnet}
diff_net <- NetCoMi::diffnet(
  net_pair,
  diffMethod = "fisher",
  adjust     = "lfdr",   # local false discovery rate
  n1 = 20, n2 = 20       # sample sizes
)

```

---

## 6  Subgraph analysis of key metabolites

The paper highlights four faecal metabolites of biological interest. Here we extract
their immediate neighbourhood from the HB igraph object to create a focused subgraph
showing which PFAM features they co-occur with.

```{r subgraph-extract}
# The four metabolites of interest
focal_metabolites <- c("fecal535neg", "fecal1214pos", "fecal367neg", "fecal703neg")

# igraph::neighborhood() returns all vertices within 1 hop of each focal node
nb <- igraph::neighborhood(
  ig_hb, order = 1,
  nodes   = igraph::V(ig_hb)[focal_metabolites],
  mindist = 0
)
# Take the union of all those neighbourhoods
union_vertices <- Reduce(union, nb)
subg <- igraph::subgraph(ig_hb, union_vertices)

cat("Subgraph: ", igraph::vcount(subg), "vertices,",
    igraph::ecount(subg), "edges\n")
```

```{r subgraph-filter}
# Keep only:
#   inter-domain edges (PFAM ↔ metabolite)  OR
#   metabolite–metabolite edges
# This removes PFAM–PFAM edges that are not relevant to the focal metabolites

# Metabolite-only subgraph to get metabolite–metabolite edges
subg_metab <- igraph::subgraph(
  subg, igraph::V(subg)[igraph::V(subg)$type == "b"]
)
metab_edge_ids <- igraph::get.edge.ids(
  subg,
  t(igraph::as_edgelist(subg_metab, names = TRUE))
)

# Inter-domain edges in the full subgraph
inter_edge_ids <- which(igraph::E(subg)$inter)

keep_edges <- union(metab_edge_ids[metab_edge_ids > 0], inter_edge_ids)
subg_filt  <- igraph::subgraph.edges(subg, keep_edges)

cat("Filtered subgraph: ", igraph::vcount(subg_filt), "vertices,",
    igraph::ecount(subg_filt), "edges\n")
```

```{r subgraph-plot, fig.width=7, fig.height=7}
# Node colour: focal metabolites one colour, neighbours coloured by type
node_col <- case_when(
  igraph::V(subg_filt)$name %in% focal_metabolites ~ "gold",
  igraph::V(subg_filt)$type == "a"                 ~ "#f58231",
  igraph::E(subg_filt)$corr > 0 ~ unname(BIOACTIVE_COLORS["High Bioactive"]),
  TRUE ~ "grey70"
)
# One colour per vertex based on its own enrichment direction
vert_cols <- case_when(
  igraph::V(subg_filt)$name %in% focal_metabolites ~ "gold",
  igraph::V(subg_filt)$type == "a"                 ~ "#f58231",
  TRUE                                              ~ "grey70"
)

# Label focal metabolites prominently; label other metabolites by annotation;
# suppress labels on PFAM nodes (too many)
vert_labels <- igraph::V(subg_filt)$name |> map_chr(\(nm) {
  if (nm %in% focal_metabolites) return(nm)
  if (igraph::V(subg_filt)[nm]$type == "b") {
    lbl <- tbl_node_md$plot_manual_label[tbl_node_md$node_id == nm]
    if (length(lbl) > 0 && !is.na(lbl[1])) return(lbl[1])
  }
  return("")
})

# Layout that keeps positively-correlated neighbours closer together
set.seed(42)
lay <- igraph::layout_with_graphopt(subg_filt)

plot(
  subg_filt,
  layout            = lay,
  vertex.size       = ifelse(igraph::V(subg_filt)$name %in% focal_metabolites, 8, 4),
  vertex.color      = vert_cols,
  vertex.shape      = igraph::V(subg_filt)$shape,
  vertex.label      = vert_labels,
  vertex.label.color = "black",
  vertex.label.family = "sans",
  vertex.label.cex  = ifelse(
    igraph::V(subg_filt)$name %in% focal_metabolites, 0.7, 0.45
  ),
  vertex.label.dist = 0.8,
  vertex.label.degree = -pi/2,
  edge.color        = igraph::E(subg_filt)$color,   # green = pos, red = neg
  edge.width        = log10(abs(igraph::E(subg_filt)$corr) * 200 + 1),
  main = "Neighbourhood subgraph of focal metabolites (HB network)\nGold = focal metabolites · Orange = PFAM · Green edges = positive"
)
```

---

## Recap

| Term | Function | What it does |
|---|---|---|
| SPIEC-EASI | `spiec.easi()` | Infers sparse partial-correlation network robust to compositionality |
| Degree | `igraph::degree()` | Number of edges per node |
| Betweenness | `igraph::betweenness()` | Fraction of shortest paths passing through a node |
| Network construction | `NetCoMi::netConstruct()` | Prepares association matrices for paired NetCoMi analysis |
| Network analysis | `NetCoMi::netAnalyze()` | Computes centrality, identifies hubs, detects communities |
| Network comparison | `NetCoMi::netCompare()` | Tests whether global properties differ between two networks |
| Differential edges | `NetCoMi::diffnet()` | Fisher test on individual edges that change between conditions |
| Neighbourhood | `igraph::neighborhood()` | Vertices within n hops of focal nodes |
| Subgraph | `igraph::subgraph.edges()` | Extract a subgraph keeping only specified edges |

---